---
title: "Data Science for Biologists Final Project: EDA"
author: "Jack Davies u7921974"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

## Location on Github

https://github.com/anu-jackdavies/DS4B-final-project


#### My Eucalypt Species - Eucalyptus decipiens

# Download ALA data and IUCN map

```{r}
library(galah)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(readxl)
library(janitor)
library(purrr)
library(sf)  
library(ozmaps)
library(here) 
```

Using galah_call to get ALA occurrence data for **Eucalyptus decipiens**:
```{r}
galah_config(email = "iamjackdavies@gmail.com")
galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_counts()
```

```{r}
galah_config(email = "iamjackdavies@gmail.com")
decipiens <- galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_occurrences()
```
Importing IUCN expert range map data for E. decipiens:
```{r}
IUCN_map <- st_read("raw_data/Eucalyptus_decipiens.gpkg")
```
Now 'IUCN_map' can be used to add the spatial polygon 

# Clean occurrence data

This will be a key part of the project which needs much consideration.

Firstly, I'll plot all of the occurrence data onto a map, then overlay the expert range.
This will be a good way to visualise what the spatial distribution of E. decipiens looks like.

## Plot ALA occurrence data

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3')
```
Could potentially flip the point with latitude around +33 since if it were -33 it would align with many other points - look into this.


### Overlapping observations

Trying to visualise overlapping observations like in Week 9 workshop:
```{r}
library(ggpointdensity)
library(viridis)
```

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```
Two key clusters are present in and around Perth (largest group, likely biased to some extent due to population), and in the southeast corner of the expert range which is overlaid in the next plot.


## Overlay IUCN expert range on occurrence map
```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F")
```

### Remove NA coordinates

```{r}
sum(is.na(decipiens$decimalLatitude))
sum(is.na(decipiens$decimalLongitude))
sum(is.na(decipiens$decimalLatitude) & is.na(decipiens$decimalLongitude))
```
146 of the 697:
```{r}
146/697
```
So about 21% of the occurrence data have no latitude/longitude coordinates and can't be used for subsequent analysis.
```{r}
697 - 146
```
This leaves 551 records, however the handout sheet said 453, strange.
Removing NA coordinate observations:

```{r}
decipiens_clean <- decipiens %>% 
  filter(!is.na(decimalLatitude),
                !is.na(decimalLongitude)
  )
```

'decipiens_clean' has 551 observations, so the code above worked.

Perhaps there have been more observations added to the dataset since the paper handout was made (551 vs 453 records on my sheet)?

### Investigate obvious spatial outliers.

Firstly, I want to look at the point with the northern latitude since it seems like it could be a valid observation, just with error caused by entering latitude with + instead of -.

```{r}
#View(decipiens_clean)
```

When I ran this and sorted by latitude I found the observation had latitude 31.92, while many other observations had -35, -34, -33, -32, -31, -30 etc, so my theory seems to hold up.
I found it was from the Royal Botanic Gardens, Kew - Herbarium Specimens. That's crazy, since that's a location in Southwest London! I noted it was a preserved specimen rather than a human observation, so perhaps it was brought to London from Australia, and its Australian location was shoddily recorded with incorrect latitude?

On ALA, I found that there are many preserved specimen records whose data resource is the Royal Botanic Gardens, Kew.

I want to check these out in more detail to see if their coordinate data is reputable, to determine if they should be kept and if I should flip that one point.

```{r}
kew_decipiens <- decipiens_clean %>% 
  filter(dataResourceName == 'Royal Botanic Gardens, Kew - Herbarium Specimens')
```

There are only 4 entries here but around 20 on ALA, meaning most of the occurrences from Kew had no coordinate information. Check out coordinates of the four entries:

```{r}
kew_decipiens$decimalLatitude
kew_decipiens$decimalLongitude
```
These seem to line up well with the other observations on the map and fall in and around expert range. Following my investigation, it seems reasonable to accept the location data for these records and correct the north latitude point to be -31.92 instead.

Flipping coordinate:
```{r}
decipiens_clean <- decipiens_clean %>% 
  mutate(decimalLatitude = if_else(decimalLatitude > 0, -decimalLatitude, decimalLatitude))
```

Note in this case that was the only north latitude point so I don't need to worry about the code incorrectly flipping other points' coordinates.

Remake the map to check:

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F")
```
Success!

Next up I want to investigate the observations on Australia's East coast which seem very out of place.

Looking at observations with high longitude, they are all super old (1960-1971) records from AD AVH Data and Canberra. They are also preserved specimens. 

### Removing Preserved Specimens

After chatting with Rob, I've confirmed that preserved specimens should be removed since they're not a true indication of where E. decipiens naturally occur.

This makes it easier, as many of the problematic-looking points seem to be preserved specimens.
```{r}
nrow(decipiens_clean)
```

```{r}
decipiens_clean <- decipiens_clean %>% 
  filter(!basisOfRecord == 'PRESERVED_SPECIMEN')
```

```{r}
nrow(decipiens_clean)
```
Wow! That removed a whopping
```{r}
551-185
```
observations! Now viewing the map again:

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.4)
```

Now the IUCN expert range is starting to look a lot more apt than it did before!
Let's zoom in to the relevant area a bit more.
```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.4) +
  coord_sf(xlim = c(114, 120),
           ylim = c(-36, -29),
           expand = FALSE)
# Zoom code made with help of ChatGPT:
# https://chatgpt.com/share/6a066eec-e304-83ec-b523-cf49140ca891
```

Note that in the top tip of the IUCN expert range, there was a larger cluster of points around the 2 observations that remain there in my cleaned dataset. Was there some reason for the IUCN to include more of them? Well, looking at the full dataset prior to cleaning, almost all points in that area (around 30S, 116E) were preserved specimens, except the exact 2 that I have left. For this reason I'll stand by my removal decision.

I'm now going to remake the ggpointdensity plot to see how well overlapping observations have been dealt with after some cleaning:

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) + 
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.2) +
  coord_sf(xlim = c(114, 120),
           ylim = c(-36, -29),
           expand = FALSE) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```

Looking on Google Maps, the large cluster is indeed around Perth, especially north of Perth where there is more forest and less urban sprawl. For example the Gangara-Moore River State forest has coords around -31.5, 116 which lines up with high clustering of E. decipiens occurrence records. 

https://www.google.com/maps/@-31.5593691,115.8296231,25841m/data=!3m1!1e3?entry=ttu&g_ep=EgoyMDI2MDUxMi4wIKXMDSoASAFQAw%3D%3D


I also noticed the other, less prominent cluster has many points in and around Stirling National Park.

https://www.google.com/maps/@-34.3899856,117.3228533,196844m/data=!3m1!1e3?entry=ttu&g_ep=EgoyMDI2MDUxMi4wIKXMDSoASAFQAw%3D%3D

It is my suspicion that there are far more E. decipiens out there in the other forest areas that are less visited by humans, and these clusters are primarily the result of bias towards more populated/visited areas, parks and roads.

Some of the entries have NA occurrence date. Since later on I will be projecting the SDM into the future (2100), I can't rely on undated occurrences to build or project the model. This is because if I don't know what date the occurrence is from, how can I know how it's position might change over time?

Removing undated entries:
```{r}
decipiens_clean <- decipiens_clean %>% 
  filter(!is.na(eventDate))
```

Now there are only 161 entries. Replotting:
```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) + 
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.2) +
  coord_sf(xlim = c(114, 120),
           ylim = c(-36, -29),
           expand = FALSE) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```

### Investigate last outliers

Before fitting my SDM I want to investigate the two easternmost points and the two northernmost points since they are either far out of the expert range or otherwise distant from other occurrences.

The easternmost point is from PlantBank Records and recorded by Donald F. Blaxell in 1986, a reputable botanist with an interest in Eucalyptus. 
https://www.anbg.gov.au/biography/blaxell-donald-frederick.html

The ALA page for the occurrence also shows no obvious signs of irregularity compared to other E. decipiens observations so I'll keep this observation.
https://biocache.ala.org.au/occurrences/72b4f887-718f-433b-ba49-865a90d017eb

The other eastern point is from the Salinity Action Plan Flora Survey in 1999.
It has good info available on the ALA page and it was recorded by members of the Department of Conservation and Land Management (M N Lyons and S D Lyons), so I will keep it in the dataset too.
https://biocache.ala.org.au/occurrences/d63a3a75-e955-412e-83b6-302a712695aa
https://researchdata.edu.au/salinity-action-plan-flora-survey/3877978


Issue with Eastern points! French et. al (2012) stated that E. decipiens intergrades with another species E. obesa and then that intergrades with E. adesmophloia at the northern slopes of the Stirling range and the eastern-adjacent plain respectively. This is shown with a map that makes it likely that the older (1986 and 1999) observations are likely from one of these other related species and were mislabelled as E. decipiens prior to the knowledge by French et. al (2012).
https://florabase.dbca.wa.gov.au/science/nuytsia/653.pdf (see page 37)

I have retroactively decided to remove these observations due to this discovery of potential species misidentification.

```{r}
decipiens_clean <- decipiens_clean %>% 
  filter(decimalLongitude < 119)
```


Northernmost point:
Collected then recorded on iNaturalist by Dean Nicolle in 2010, a leading expert in eucalyptus botany. It has plenty of information on ALA including that it is wild, not cultivated. For these reasons the data point is reputable and it will be included. The fact it was by Nicolle may have influenced the IUCN experts to make their range wrap around this point too.
https://biocache.ala.org.au/occurrences/fc615d1b-3c6f-48fb-9a9c-0440304758c1
https://www.anbg.gov.au/biography/nicolle-dean.html

Second northern point:
Identified by Dean Nicolle again in 2024 using iNaturalist, collector/user is Malcolm Simard, a non-professional but likely accompanied by Nicolle. The observation is labelled as research grade by ALA and was overseen by an expert so I will include it in my data.
https://biocache.ala.org.au/occurrences/a47f74ac-9bdc-4987-b933-b6b8a6bef9a3


### Final check with coordinate cleaner:

This is a useful tool from Dax's book to identify any further points for removal.
```{r}
library(CoordinateCleaner)

# Run record-level tests
coordinate_tests <- clean_coordinates(x = decipiens_clean, 
                                      species = "scientificName")
summary(coordinate_tests)
```

None of the data which has survived my cleaning was flagged as suspicious by coordinate cleaner, which is a good sign that this data has been well cleaned and is ready for modelling. 


# Assess whether expert range map should be changed

Considering the data cleaning I have performed, I do believe the IUCN expert range map should be slightly modified.

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) + 
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.2) +
  coord_sf(xlim = c(114, 120),
           ylim = c(-36, -29),
           expand = FALSE) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```


However, the extent to which it should be changed is dependent on interpretation, the approaches I am considering is:

a) Adjust the range to include all of the clean data points i.e. more east to cover Stirling Range, cover second northern point.

b) Make only slight shifts to the expert range map to better reflect the clean data I've chosen to use, but not having to cover every single occurrence point e.g. cutting certain parts inwards to remove blank space without observations in my data and not forcing range to go as far east or cover either of the northernmost points if the model decides against it.

The process of building an SDM will help validate/change these approach ideas.

*important note: The data which I initially loaded into R, 'decipiens', had ALA filters already applied to it e.g. only including 'wild' labelled occurrences, not 'cultivated'. So other concerns I haven't mentioned here were likely already addressed by the ALA filters.

# Build SDM for E. decipiens suitable climate space

## Background Points

Before loading variables and model selection, it is important to choose background points.

Of the four main strategies for choosing background points, I believe that accessible area is the clear candidate for E. decipiens. This is because as with most/all eucalypts, they are unlikely to disperse very far at all, with Booth (2017) presenting a typical seed dispersal limit of around 5km until 2085 across over 600 eucalypts.
https://doi.org/10.1071/BT17019

Environmental envelope would ignore the incapability of far-reaching dispersal for E. decipiens, while the other two methods sample across all of Australia which would not be appropriate for this data which is localised to the corner of Western Australia. Hence, I will be using the accessible area background points approach.

First I need to load the bioclim raster which contains the 19 bioclim variables from CHELSA v2.1 BIOCLIM, as 'bioclim_data' is needed in the code chunk.

```{r}
library(raster) 
library(terra)
library(skimr)
```

```{r}
bioclim_data <- rast("raw_data/worldclim_australia.tif")
```

Pruning data to one point per grid cell, using Alex's week 8 workshop code:
```{r}
# prune data to one point per grid cell
# Create SpatVector
occ_v <- vect(decipiens_clean, geom = c("decimalLongitude", "decimalLatitude"), crs = crs(bioclim_data))

# Get raster cell IDs
occ_v$cell_id <- cellFromXY(bioclim_data, crds(occ_v))

# Keep one point per raster cell
occ_pruned <- occ_v[!duplicated(occ_v$cell_id), ]

# Extract climate
occ_env <- terra::extract(
  bioclim_data,
  occ_pruned,
  ID = FALSE
) |> 
  as_tibble()

occ_model <- bind_cols(
  as_tibble(crds(occ_pruned)) |> 
    setNames(c("longitude", "latitude")),
  occ_env
) |> 
  mutate(presence = 1)
```


Now I can use more code from the week 8 workshop to make the accessible area background points.

```{r}
# Convert occurrence data to sf object (needed for spatial operations)
occ_sf <- st_as_sf(decipiens_clean, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

occ_proj <- st_transform(occ_sf, 3577)

buffer <- st_buffer(occ_proj, dist = 50000) |>
  st_union()

buffer_vect <- vect(st_transform(buffer, 4326))

bg <- spatSample(
  mask(bioclim_data, buffer_vect),
  size = 5000,
  method = "random",
  na.rm = TRUE,
  as.points = TRUE
)

bg_env <- terra::extract(bioclim_data, bg, ID = FALSE) |>
  as_tibble()

bg_model <- bind_cols(
  st_coordinates(st_as_sf(bg)) |>
    as_tibble() |>
    rename(longitude = X, latitude = Y),
  bg_env
) |>
  mutate(presence = 0)
```

I chose to implement a 50km buffer somewhat arbitrarily, as an approximation or guess of where more unobserved E. decipiens occurrences could feasibly be without being recorded yet. I tested larger and smaller values, but too large meant the background points covered the whole southwestern corner of WA (not terrible but unnecessary) and too small left gaps between known observations which I think could contain legitimate observations.

### Plotting Background points on map

Using leaflet, I'll quickly visualise the background points to make sure it's working before model fitting.
Again, I'll be utilising the week 8 workshop code:

```{r}
library(leaflet)
```

```{r}
leaflet() |>
  addProviderTiles(providers$OpenStreetMap.Mapnik) |>

  # Add a climate raster layer (e.g. Bio01)
  addRasterImage(raster::raster(bioclim_data$Bio01), opacity = 0.5) |>

  # Plot background points in grey
  addCircleMarkers(
    data = bg_model,
    lng = ~longitude, lat = ~latitude,
    radius = 3,
    fillOpacity = 0.4,
    stroke = FALSE,
    color = "grey"
  ) |>

  # Plot occurrence points in blue on top
  addCircleMarkers(
    data = occ_model,
    lng = ~longitude, lat = ~latitude,
    radius = 5,
    fillOpacity = 0.7,
    stroke = FALSE,
    color = "#2C7FB8"
  )
```


### Load CHELSA V2.1 historical bioclim variables

The current 'bioclim_data' has a low resolution:
```{r}
res(bioclim_data)
# about 9km according to Claude https://claude.ai/share/ec09d26e-7d80-4b1b-aa6f-923c79ade46a
```
and seems to be from worldclim not CHELSA V2.1 bioclim. Because of this I'll make sure to import the higher res CHELSA bioclim variables now:


```{r}
if (!file.exists("raw_data/chelsa_bioclim_cropped.tif")) {
  bioclim_chelsa <- lapply(1:19, function(i) {
    url <- paste0("/vsicurl/https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio",
                  i, "_1981-2010_V.2.1.tif")
    message("Loading and cropping bio", i, "...")
    crop(rast(url), buffer_vect)
  }) |> rast()
  writeRaster(bioclim_chelsa, "raw_data/chelsa_bioclim_cropped.tif", overwrite = TRUE)
}

bioclim_chelsa <- rast("raw_data/chelsa_bioclim_cropped.tif")
# made with Claude asssistance to load variables without requiring physical download
# https://claude.ai/share/ec09d26e-7d80-4b1b-aa6f-923c79ade46a
```

The above chunk may take a long time to load the data, so I've used the if condition such that if you download 'chelsa_bioclim_cropped.tif', you can skip the long part saving time on knitting.

```{r}
res(bioclim_chelsa)
```
This is 10x better resolution (30 arcseconds, around 1km resolution). The reason I went through this is because the soil variables used below have 90m resolution, so having to downscale it less will be better for model fitting (aka I've tried to minimise gap in resolutions so scaling down soil variable res won't be amplifying as much noise)

### Load useful soil variables

Soil is a key area we've been instructed to investigate, with some recommended sources.

This page is where the data can be looked at or accessed manually if you wish: https://esoil.io/TERNLandscapes/Public/Pages/SLGA/GetData-COGSDataStore.html

E. decipiens grows in clay, loam or sandy soils, and grow over laterite.
https://www.inaturalist.org/taxa/602566-Eucalyptus-decipiens

Studies have shown that lateritic soils tend to have low soil organic carbon (SOC), which is a useful soil variable from SLGA that I can include (Sow and Kundu, 2020).
https://doi.org/10.22271/chemi.2020.v8.i3al.9611

Considering my research and the available variables, I'll include clay, soil organic carbon, and soil ph. These may be reduced later in model selection.

Importing the soil variables from soilworld:
```{r}
library(geodata)
library(terra)

geodata_path("geodata")

clay <- soil_world(var = "clay", depth = 30, stat = "mean")
soc  <- soil_world(var = "soc", depth = 30, stat = "mean")
ph   <- soil_world(var = "phh2o", depth = 30, stat = "mean")

# Assisted with ChatGPT - scroll to very bottom for this, since trying to use SLGA before was catastrophic and lengthy: https://chatgpt.com/share/6a0fe8c8-bd28-83ec-8bd8-ea7dfd13b5c0
```


Cropping soil variable data to the buffer area from background points:
```{r}
clay_crop           <- crop(clay,    buffer_vect)
soc_crop    <- crop(soc,    buffer_vect)
ph_crop <- crop(ph, buffer_vect)
# done with Claude help: https://claude.ai/share/ec09d26e-7d80-4b1b-aa6f-923c79ade46a
```


Getting the soil variables to match CHELSA bioclim resolution:
```{r}
clay_crop           <- resample(clay_crop,    bioclim_chelsa)
soc_crop    <- resample(soc_crop,    bioclim_chelsa)
ph_crop <- resample(ph_crop, bioclim_chelsa)

# done with Claude help: https://claude.ai/share/ec09d26e-7d80-4b1b-aa6f-923c79ade46a
```


### Model Selection

I'll adapt the code I used in the Week 8 workshop to filter out bioclim variables with strong correlations to others. Since the variables are all either rainfall or temperature related, there is plenty of correlation so many should be removed:

```{r}
sdm_data <- bind_rows(occ_model, bg_model)
```

Checking for bioclim collinearity:
```{r}
library(caret)
```

```{r}
bio_cols <- grep("^Bio", colnames(sdm_data), value = TRUE)
correlation_matrix <- cor(sdm_data[, bio_cols], use = "complete.obs")

correlated <- findCorrelation(correlation_matrix, cutoff = 0.7)  # |r| > 0.7
names_correlated <- bio_cols[correlated]
```

```{r}
# Remove highly correlated variables
sdm_data_uncor <- sdm_data[, !colnames(sdm_data) %in% names_correlated]

# Keep track of predictors
predictor_vars <- grep("^Bio", colnames(sdm_data_uncor), value = TRUE)
print(predictor_vars)
```
Mean Daily Mean Near-Surface Air Temperature of the Quarter (bio11),
Mean Monthly Precipitation of the Warmest Quarter (bio18),
Mean Monthly Precipitation of the Coldest Quarter (bio19),
Isothermality (bio03),
Annual Daily Mean Near-Surface Air Temperature Range (bio07) remain.

I want to include a temperature variable, not just isothermality or range, and while each of the max/min/mean avg temperature Bio variables are highly correlated with each other, I can still justifiably include *one* of these variables. I'll choose to include bio05, the Mean Daily Maximum Near-Surface Air Temperature of the Warmest Month as it will test how E. decipiens can cope with hot desert climates near its current distribution, and will be good for assessing future ranges as temperatures increase.

```{r}
predictor_vars <- c(predictor_vars, names(sdm_data["Bio05"]))
sdm_data_uncor <- c(sdm_data_uncor, names(sdm_data["Bio05"]))
```


Now I'll group in the soil variables:
```{r}
selected_predictors <- c(
  bioclim_chelsa[[c("CHELSA_bio11_1981-2010_V.2.1",
                    "CHELSA_bio18_1981-2010_V.2.1",
                    "CHELSA_bio19_1981-2010_V.2.1",
                    "CHELSA_bio3_1981-2010_V.2.1",
                    "CHELSA_bio7_1981-2010_V.2.1",
                    "CHELSA_bio5_1981-2010_V.2.1")]],
  clay_crop,
  soc_crop,
  ph_crop
)

names(selected_predictors) <- c("Bio11", "Bio18", "Bio19", "Bio03", "Bio07", "Bio05", "Clay", "SOC", "pH")
# done with Claude help: https://claude.ai/share/ec09d26e-7d80-4b1b-aa6f-923c79ade46a
```

Simplified names of predictors for ease of use.

I need to re-extract and rebuild sdm_data with the soil variables now:
```{r}
occ_env <- terra::extract(
  selected_predictors,
  occ_pruned,
  ID = FALSE
) |> 
  as_tibble()

occ_model <- bind_cols(
  as_tibble(crds(occ_pruned)) |> 
    setNames(c("longitude", "latitude")),
  occ_env
) |> 
  mutate(presence = 1)

bg_env <- terra::extract(selected_predictors, bg, ID = FALSE) |>
  as_tibble()

bg_model <- bind_cols(
  st_coordinates(st_as_sf(bg)) |>
    as_tibble() |>
    rename(longitude = X, latitude = Y),
  bg_env
) |>
  mutate(presence = 0)

sdm_data <- bind_rows(occ_model, bg_model)
# Advice on how to tweak code here from Claude https://claude.ai/share/ec09d26e-7d80-4b1b-aa6f-923c79ade46a
```

Update predictors:
```{r}
predictor_vars <- c("Bio11", "Bio18", "Bio19", "Bio03", "Bio07", "Bio05", "Clay", "SOC", "pH")
```

## Fitting GLM and GAM

Following the week 8 workshop process:
```{r}
# GLM: linear model
formula_glm <- formula(paste0("presence ~ ", paste0(predictor_vars, collapse = " + ")))

# GAM: non-linear model with splines (k = basis dimension, controls wiggliness)
formula_gam <- formula(paste0("presence ~ ", paste0("s(", predictor_vars,", k = 4)", collapse = " + ")))

# Inspect formulas
print(formula_glm)

print(formula_gam)
```

Using the selected predictors in a GLM:
```{r}
# Fit GLM (logistic regression)
glm_sdm <- glm(formula_glm, data = sdm_data, family = binomial(link = "logit"))
summary(glm_sdm)
```

```{r}
# Fit GAM (using mgcv, which correctly handles smooth terms in predict)
gam_sdm <- mgcv::gam(formula_gam, data = sdm_data, family = binomial(link = "logit"))
summary(gam_sdm)
```

```{r}
gam_sdm$aic
```

Looking at potential variable removal, the GLM suggests many variables aren't significant, with particular doubts on Bio11 (Mean Daily Mean Near-Surface Air Temperature of the Quarter), Bio07 (Annual Daily Mean Near-Surface Air Temperature Range), Bio05 (Mean Daily Maximum Near-Surface Air Temperature of the Warmest Month) and soil pH.

The GAM summary changes the overall picture. Variables which are still have p-values well over 0.05 are strong candidates for removal, these being Bio07, pH. Some of the other doubtful variables are far more significant than in GLM like Bio05 and Bio11, suggesting they do explain variation in presence to a significant level, just non-linearly. Clay is far closer to the 0.05 cutoff under GAM than in GLM so I won't remove it yet, while SOC is actually less significant under the GAM, indicating there could be a linear association where more SOC means greater habitat suitability without non-linear nuances.

Note that GAM has lower AIC and better deviance explained than GLM, indicating better model fit. Also note I tried fitting GAM with different k-values but k = 4 worked the best (had more significant variables).

So, I'll remove Bio07, pH. I also will remove Bio18 since Bio18 and Bio19 are both precipitation measures just in different quarters, and of those two Bio19 was more significant.

```{r}
sdm_data <- dplyr::select(sdm_data, -Bio07, -pH)
predictor_vars <- predictor_vars[!predictor_vars %in% c("Bio07", "pH")]

# made with Claude help https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```

```{r}
# GLM: linear model
formula_glm <- formula(paste0("presence ~ ", paste0(predictor_vars, collapse = " + ")))

# GAM: non-linear model with splines (k = basis dimension, controls wiggliness)
formula_gam <- formula(paste0("presence ~ ", paste0("s(", predictor_vars,", k = 4)", collapse = " + ")))

# Inspect formulas
print(formula_glm)

print(formula_gam)
```

```{r}
# Fit GLM (logistic regression)
glm_sdm <- glm(formula_glm, data = sdm_data, family = binomial(link = "logit"))
summary(glm_sdm)
```

```{r}
# Fit GAM (using mgcv, which correctly handles smooth terms in predict)
gam_sdm <- mgcv::gam(formula_gam, data = sdm_data, family = binomial(link = "logit"))
summary(gam_sdm)
```
The deviance explained and adjusted R^2 have ever so slightly reduced and AIC increased after dropping Bio07 and pH but this is not of major concern due to the benefit of a more simple model for the SDM. Note I considered dropping Bio18 since 18 and 19 are both precipitation variables and Bio19 was more significant, but actually this model does better than one without Bio18 and the two of those variables have a significant impact on presence in their own right whilst passing the 0.7 correlation cutoff earlier.

While Clay and SOC are not statistically significant model inclusions they aren't far from it, and I would like to include soil components into the SDM as I believe it to be an important biological aspect to consider for habitat suitability.

Update with removed predictors:
```{r}
selected_predictors <- selected_predictors[[names(selected_predictors)[!names(selected_predictors) %in% c("Bio07", "pH")]]]

# Made with Claude assistance: https://claude.ai/share/f4531869-870a-4c7f-ba27-5ee61fcf997a
```

Checking out AUC as I have in Week 8:

```{r}
library(dismo)
```

```{r}
# Split presence and background, drop rows with missing predictors so
# predict() returns a clean, matching-length vector.
p <- na.omit(sdm_data[sdm_data$presence == 1, ])
a <- na.omit(sdm_data[sdm_data$presence == 0, ])

# Get predicted suitability for presence and background points separately, and
# strip dims/names with as.numeric() so dismo::evaluate() recognises them as
# plain vectors (predict.gam returns a 1-D array, which dismo misinterprets).
glm_pred_p <- as.numeric(predict(glm_sdm, newdata = p, type = "response"))
glm_pred_a <- as.numeric(predict(glm_sdm, newdata = a, type = "response"))
gam_pred_p <- as.numeric(predict(gam_sdm, newdata = p, type = "response"))
gam_pred_a <- as.numeric(predict(gam_sdm, newdata = a, type = "response"))

# Evaluate models using the predicted values directly.
# evaluate() returns an S4 object — use @ (not $) to access slots, e.g. glm_eval@auc
glm_eval <- evaluate(p = glm_pred_p, a = glm_pred_a)
gam_eval <- evaluate(p = gam_pred_p, a = gam_pred_a)

# Plot ROC curves
par(mfrow = c(1, 2))
plot(glm_eval, "ROC")
plot(gam_eval, "ROC")
```
According to Lecture 8, AUC of 1 is perfect and around 0.8 is indicative of a reasonable SDM. The GAM AUC of 0.893 is quite good therefore.

# Plot the SDM

Now I want to make plots with the GLM, GAM, expert range and alpha hull to compare them.

I can do this by adapting the leaflet plotting code I worked through in the Week 8 workshop.
I'd like to use static plots instead of leaflet as it's less complex and makes for simpler interpretation in the presentation.

```{r}
library(patchwork)
library(tidyterra)
library(rangeBuilder)
```
Generating predictions for current distribution:

```{r}
pred_glm <- terra::predict(selected_predictors, glm_sdm, type = "response")
pred_gam <- terra::predict(selected_predictors, gam_sdm, type = "response")

# Made with Claude assistance: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```


Getting 95% alpha hull:
```{r}
hull <- getDynamicAlphaHull(
  x = decipiens_clean[, c("decimalLongitude", "decimalLatitude")],
  fraction = 0.95,
  partCount = 1
)
hull_sf <- st_as_sf(hull[[1]])

# Made with Claude assistance: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```

As I'm working with geom sf and static plot not leaflet, I can't set the alpha value myself. Instead the code dynamically sets alpha such that the specific fraction of the occurrences are encompassed by the hull.

GLM Plot:
```{r}
plot_glm <- ggplot() +
  geom_spatraster(data = pred_glm) +
  scale_fill_distiller(palette = "Spectral", na.value = "transparent",
                       limits = c(0, 1), name = "Suitability") +
  geom_point(data = decipiens_clean, aes(x = decimalLongitude, y = decimalLatitude),
             colour = "pink", size = 0.8, alpha = 0.5) +
  labs(title = "GLM Suitability") +
  theme_minimal()
# Made with Claude assistance: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```

GAM plot:
```{r}
plot_gam <- ggplot() +
  geom_spatraster(data = pred_gam) +
  scale_fill_distiller(palette = "Spectral", na.value = "transparent",
                       limits = c(0, 1), name = "Suitability") +
  geom_point(data = decipiens_clean, aes(x = decimalLongitude, y = decimalLatitude),
             colour = "pink", size = 0.8, alpha = 0.5) +
  labs(title = "GAM Suitability") +
  theme_minimal()
# Made with Claude assistance: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```

Alpha hull + expert range plot:
```{r}
p3 <- ggplot() +
  geom_sf(data = IUCN_map, fill = "lightgreen", colour = "darkgreen",
          alpha = 0.4, linewidth = 0.5) +
  geom_sf(data = hull_sf, fill = "steelblue", colour = "steelblue",
          alpha = 0.2, linewidth = 0.5, linetype = "dashed") +
  geom_point(data = decipiens_clean, aes(x = decimalLongitude, y = decimalLatitude),
             colour = "pink", size = 0.8, alpha = 0.5) +
  labs(title = "Expert Range vs Alpha Hull") +
  theme_minimal()
# Made with Claude assistance: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```


Combine all of the plots:
```{r}
plot_glm + plot_gam + p3 +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Eucalyptus decipiens — Habitat Suitability & Range Comparison",
    theme = theme(plot.title = element_text(hjust = 0.5))
  )
# Made with Claude assistance: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```
### Best Model Plot

```{r}
ggplot() +
  geom_spatraster(data = pred_gam) +
  scale_fill_distiller(palette = "Spectral", na.value = "transparent",
                       limits = c(0, 1), name = "Suitability") +
  geom_sf(data = IUCN_map, aes(colour = "IUCN Range"), fill = "lightgreen",
          alpha = 0.2, linewidth = 0.7) +
  geom_sf(data = hull_sf, aes(colour = "Alpha Hull"), fill = NA,
          linewidth = 0.7) +
  geom_point(data = decipiens_clean, aes(x = decimalLongitude, y = decimalLatitude,
             colour = "Occurrences"), size = 0.9, alpha = 1) +
  scale_colour_manual(
    name = NULL,
    values = c("IUCN Range" = "darkgreen",
               "Alpha Hull" = "aquamarine",
               "Occurrences" = "hotpink")
  ) +
  labs(title = "E. decipiens GAM Suitability, IUCN Range & Alpha Hull",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()
# Made with Claude assistance, but primarily edited by me: https://claude.ai/share/1ce5afbe-ff4e-49c1-9662-a900a2581fe5
```
Note: The reason for some white spots on the map is likely due to NA entries in the rasters for clay and SOC %. 


### Thresholding suitability

As done in week 8 workshop:
```{r}
pred_glm_threshold <- threshold(glm_eval, stat = "spec_sens")
pred_gam_threshold <- threshold(gam_eval, stat = "spec_sens")

pred_glm_bin <- pred_glm > pred_glm_threshold
pred_gam_bin <- pred_gam > pred_gam_threshold
```

Making GAM vs hull vs expert range plot again with thresholding:
```{r}
ggplot() +
  geom_spatraster(data = pred_gam_bin) +
  scale_fill_manual(
    values = c("FALSE" = "#FFFDD6", "TRUE" = "#4DAF4A"),
    na.value = "transparent",
    na.translate = FALSE,
    name = "Suitability",
    labels = c("FALSE" = "Unsuitable", "TRUE" = "Suitable")
  ) +
  geom_sf(data = IUCN_map, aes(colour = "IUCN Range"), fill = "lightgreen",
          alpha = 0.2, linewidth = 0.73) +
  geom_sf(data = hull_sf, aes(colour = "95% Alpha Hull"), fill = NA,
          linewidth = 0.7) +
  geom_point(data = decipiens_clean, aes(x = decimalLongitude, y = decimalLatitude,
             colour = "Occurrences"), size = 0.9, alpha = 1) +
  scale_colour_manual(
    name = NULL,
    values = c("IUCN Range" = "darkgreen",
               "95% Alpha Hull" = "cyan2",
               "Occurrences" = "blue")
  ) +
  labs(title = "E. decipiens GAM Suitability, IUCN Range & Alpha Hull",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()

# Made with Claude assistance: https://claude.ai/share/af6e6c8c-0d2b-4f57-bf77-e20d32f3e46c
```

# Projecting Model to 2100

In order to project the model to see how E. decipiens' suitable climate is predicted to change, I need to get some CHELSA V2.1 Bioclim variables for 2071-2100.

To get the CHELSA 2071-2100 data, I need to determine which Global Climate Model (GCM) and Shared Socioeconomic Pathway (SSP) to select. 

### Select SSP and GCM

From reading Hausfather and Peters (2020), the choice of SSP depends on what I believe is a reasonable prediction of global warming up to 2100.
Considering the tracking of current policies compared to the pledged policies shown in that paper, I have determined that a weak mitigation of global warming (+3 degrees by 2100) is probably most likely, while +2.5 degrees may be possible if the progress and policies improve. I'll therefore use SSP4 - 6.0 (+3 degrees) and then also test SSP2 - 4.5 (+2.5) to see how that changes my model projection.

Hausfather et. al (2022) warns that the CMIP6 GCMs can be highly-sensitive and a "hot-model" which projects extreme temperature increase of 5+ degrees regularly due to noise/variation influencing the mean per month/year etc. The paper discusses how the IUCN's AR6 report excluded the "too-hot" models from CMIP6 to avoid over sensitivity and get a realistic projection, instead of opting for the older CMIP5. I will opt to use CMIP6 but filter to only use GCMs for the 'likely' +2.5-4 degree range to get a realistic AR6 like projection (Hausfather et. al, 2022).


### Import CHELSA GCM/SSP data

There are 5 GCM options which are all CMIP 6 models.
I will choose to use MPI-ESM1-2-HR GCM, since this has a 3 degree sensitivity which fits well with my prior discussion.
https://doi.org/10.1029/2017MS001217

Within this in CHELSA, the SSP options are only SSP1 - 2.6, SSP3 - 7.0 and SSP5 - 8.5. I'll choose SSP3 as this is between the SSP2 and SSP4 options I thought would be good before.

Getting Bioclim variables which were in my SDM model:

Bio11 (mean daily mean surface temp of coldest quarter)
```{r}
url <- "/vsicurl/https://os.unil.cloud.switch.ch/chelsa02/chelsa/global/bioclim/bio11/2071-2100/MPI-ESM1-2-HR/ssp370/CHELSA_mpi-esm1-2-hr_ssp370_bio11_2071-2100_V.2.1.tif"

bio11_fut <- rast(url)


bio11_fut_crop <- crop(bio11_fut, buffer_vect)


writeRaster(bio11_fut_crop, "raw_data/bio11_australia_2071-2100.tif", overwrite = TRUE)

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```

Bio18 (mean monthly precip in warmest quarter)
```{r}
url <- "/vsicurl/https://os.unil.cloud.switch.ch/chelsa02/chelsa/global/bioclim/bio18/2071-2100/MPI-ESM1-2-HR/ssp370/CHELSA_mpi-esm1-2-hr_ssp370_bio18_2071-2100_V.2.1.tif"

bio18_fut <- rast(url)


bio18_fut_crop <- crop(bio18_fut, buffer_vect)


writeRaster(bio18_fut_crop, "raw_data/bio18_australia_2071-2100.tif", overwrite = TRUE)

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```


Bio19 (mean monthly precip in coldest quarter)
```{r}

url <- "/vsicurl/https://os.unil.cloud.switch.ch/chelsa02/chelsa/global/bioclim/bio19/2071-2100/MPI-ESM1-2-HR/ssp370/CHELSA_mpi-esm1-2-hr_ssp370_bio19_2071-2100_V.2.1.tif"

bio19_fut <- rast(url)

bio19_fut_crop <- crop(bio19_fut, buffer_vect)

writeRaster(bio19_fut_crop, "raw_data/bio19_australia_2071-2100.tif", overwrite = TRUE)

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```

Bio03 (isothermality)
```{r}

url <- "/vsicurl/https://os.unil.cloud.switch.ch/chelsa02/chelsa/global/bioclim/bio03/2071-2100/MPI-ESM1-2-HR/ssp370/CHELSA_mpi-esm1-2-hr_ssp370_bio03_2071-2100_V.2.1.tif"

bio3_fut <- rast(url)


bio03_fut_crop <- crop(bio3_fut, buffer_vect)


writeRaster(bio03_fut_crop, "raw_data/bio03_australia_2071-2100.tif", overwrite = TRUE)

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```


Bio05 (mean daily max air temp of warmest month)
```{r}

url <- "/vsicurl/https://os.unil.cloud.switch.ch/chelsa02/chelsa/global/bioclim/bio05/2071-2100/MPI-ESM1-2-HR/ssp370/CHELSA_mpi-esm1-2-hr_ssp370_bio05_2071-2100_V.2.1.tif"

bio05_fut <- rast(url)


bio05_fut_crop <- crop(bio05_fut, buffer_vect)


writeRaster(bio05_fut_crop, "raw_data/bio05_australia_2071-2100.tif", overwrite = TRUE)

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```


Stack the future Bio variables into a single raster
```{r}
future_stack <- c(bio11_fut_crop,
                  bio18_fut_crop,
                  bio19_fut_crop,
                  bio03_fut_crop,
                  bio05_fut_crop,
                  clay_crop,
                  soc_crop
                  )
```

Note that only the CHELSA V2.1 BIOCLIM variables have 2100 predicted data so clay, soc will have to be extrapolated based on the current data.

## Fit 2100 Prediction Model

I can use the predict() function instead of having to refit the entire SDM:
```{r}
names(future_stack) <- names(selected_predictors)

fut_pred <- predict(future_stack, gam_sdm, fun = predict, type = "response")

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```

Thresholding:
```{r}
fut_pred_bin <- fut_pred > pred_gam_threshold
```

Creating combined layer for suitability (now can make different colours on map for no suitablity, only for current, only for 2100 or both!)
```{r}
# 0 = unsuitable, 1 = current only, 2 = future only, 3 = both
combined <- pred_gam_bin + fut_pred_bin * 2  # gives 0,1,2,3
combined <- as.factor(combined)

# Made with Claude assistance https://claude.ai/share/5d4d1e1a-2419-41d5-aa1f-8f0f170b424d
```

## 2100 Prediction Plot
```{r}
ggplot() +
  geom_spatraster(data = combined) +
  scale_fill_manual(
    values = c(
      "0" = "#FFFDD6",   # unsuitable
      "1" = "#4DAF4A",   # current only
      "2" = "#E07B3F",   # future only (orange)
      "3" = "hotpink"    # both (same as current, or pick a different colour)
    ),
    na.value = "transparent",
    na.translate = FALSE,
    name = "Suitability",
    labels = c("0" = "Unsuitable", "1" = "Current only",
               "2" = "Future only (2100)", "3" = "Current & Future")
  ) +
  geom_sf(data = IUCN_map, aes(colour = "IUCN Range"), fill = "lightgreen",
          alpha = 0.2, linewidth = 0.73) +
  geom_point(data = decipiens_clean, aes(x = decimalLongitude, y = decimalLatitude,
             colour = "Occurrences"), size = 0.9, alpha = 1) +
  scale_colour_manual(
    name = NULL,
    values = c("IUCN Range" = "darkgreen",
               "Occurrences" = "blue")
  ) +
  labs(title = "E. decipiens GAM Suitability — Current vs 2100 (SSP370)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

