---
title: "Data Science for Biologists Final Project: EDA"
author: "Jack Davies u7921974"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

## Location on Github

https://github.com/anu-jackdavies/DS4B-final-project


#### My Eucalypt Species - Eucalyptus decipiens

# Download ALA data and IUCN map

```{r}
library(galah)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(readxl)
library(janitor)
library(purrr)
library(sf)  
library(ozmaps)
library(here) 
```

Using galah_call to get ALA occurrence data for **Eucalyptus decipens**:
```{r}
galah_config(email = "iamjackdavies@gmail.com")
galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_counts()
```

```{r}
galah_config(email = "iamjackdavies@gmail.com")
decipiens <- galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_occurrences()
```
Importing IUCN expert range map data for E. decipiens:
```{r}
IUCN_map <- st_read("raw_data/Eucalyptus_decipiens.gpkg")
```
Now 'IUCN_map' can be used to add the spatial polygon 

# Clean occurrence data

This will be a key part of the project which needs much consideration.

Firstly, I'll plot all of the occurrence data onto a map, then overlay the expert range.
This will be a good way to visualise what the spatial distribution of E. decipiens looks like.

## Plot ALA occurrence data

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3')
```
Could potentially flip the point with latitude around +33 since if it were -33 it would align with many other points - look into this.


### Overlapping observations

Trying to visualise overlapping observations like in Week 9 workshop:
```{r}
library(ggpointdensity)
library(viridis)
```

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```

## Overlay IUCN expert range on occurrence map
```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F")
```

Look for and remove NA coordinates:
```{r}
sum(is.na(decipiens$decimalLatitude))
sum(is.na(decipiens$decimalLongitude))
sum(is.na(decipiens$decimalLatitude) & is.na(decipiens$decimalLongitude))
```
146 of the 697:
```{r}
146/697
```
So about 21% of the occurrence data have no latitude/longitude coordinates and can't be used for subsequent analysis.
```{r}
697 - 146
```
This leaves 551 records, however the handout sheet said 453, strange.
Removing NA coordinate observations:

```{r}
decipiens_clean <- decipiens %>% 
  filter(!is.na(decimalLatitude),
                !is.na(decimalLongitude)
  )
```

'decipiens_clean' has 551 observations, so the code above worked.

Perhaps there have been more observations added to the dataset since the paper handout was made (551 vs 453 records on my sheet)?


Once I've made my decisions on how to clean the data and remade this plot, 
I can:

# Assess whether expert range map should be changed

# Build SDM for E. decipiens suitable climate space