---
title: "Data Science for Biologists Final Project: EDA"
author: "Jack Davies u7921974"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

## Location on Github

https://github.com/anu-jackdavies/DS4B-final-project


#### My Eucalypt Species - Eucalyptus decipiens

# Download ALA data and IUCN map

```{r}
library(galah)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(readxl)
library(janitor)
library(purrr)
library(sf)  
library(ozmaps)
library(here) 
```

Using galah_call to get ALA occurrence data for **Eucalyptus decipens**:
```{r}
galah_config(email = "iamjackdavies@gmail.com")
galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_counts()
```

```{r}
galah_config(email = "iamjackdavies@gmail.com")
decipiens <- galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_occurrences()
```
Importing IUCN expert range map data for E. decipiens:
```{r}
IUCN_map <- st_read("raw_data/Eucalyptus_decipiens.gpkg")
```
Now 'IUCN_map' can be used to add the spatial polygon 

# Clean occurrence data

This will be a key part of the project which needs much consideration.

Firstly, I'll plot all of the occurrence data onto a map, then overlay the expert range.
This will be a good way to visualise what the spatial distribution of E. decipiens looks like.

## Plot ALA occurrence data

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3')
```
Could potentially flip the point with latitude around +33 since if it were -33 it would align with many other points - look into this.


### Overlapping observations

Trying to visualise overlapping observations like in Week 9 workshop:
```{r}
library(ggpointdensity)
library(viridis)
```

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```
Two key clusters are present in and around Perth (largest group, likely biased to some extent due to population), and in the southeast corner of the expert range which is overlaid in the next plot.


## Overlay IUCN expert range on occurrence map
```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F")
```

### Remove NA coordinates

```{r}
sum(is.na(decipiens$decimalLatitude))
sum(is.na(decipiens$decimalLongitude))
sum(is.na(decipiens$decimalLatitude) & is.na(decipiens$decimalLongitude))
```
146 of the 697:
```{r}
146/697
```
So about 21% of the occurrence data have no latitude/longitude coordinates and can't be used for subsequent analysis.
```{r}
697 - 146
```
This leaves 551 records, however the handout sheet said 453, strange.
Removing NA coordinate observations:

```{r}
decipiens_clean <- decipiens %>% 
  filter(!is.na(decimalLatitude),
                !is.na(decimalLongitude)
  )
```

'decipiens_clean' has 551 observations, so the code above worked.

Perhaps there have been more observations added to the dataset since the paper handout was made (551 vs 453 records on my sheet)?

### Investigate obvious spatial outliers.

Firstly, I want to look at the point with the northern latitude since it seems like it could be a valid observation, just with error caused by entering latitude with + instead of -.

```{r}
#View(decipiens_clean)
```

When I ran this and sorted by latitude I found the observation had latitude 31.92, while many other observations had -35, -34, -33, -32, -31, -30 etc, so my theory seems to hold up.
I found it was from the Royal Botanic Gardens, Kew - Herbarium Specimens. That's crazy, since that's a location in Southwest London! I noted it was a preserved specimen rather than a human observation, so perhaps it was brought to London from Australia, and its Australian location was shoddily recorded with incorrect latitude?

On ALA, I found that there are many preserved specimen records whose data resource is the Royal Botanic Gardens, Kew.

I want to check these out in more detail to see if their coordinate data is reputable, to determine if they should be kept and if I should flip that one point.

```{r}
kew_decipiens <- decipiens_clean %>% 
  filter(dataResourceName == 'Royal Botanic Gardens, Kew - Herbarium Specimens')
```

There are only 4 entries here but around 20 on ALA, meaning most of the occurrences from Kew had no coordinate information. Check out coordinates of the four entries:

```{r}
kew_decipiens$decimalLatitude
kew_decipiens$decimalLongitude
```
These seem to line up well with the other observations on the map and fall in and around expert range. Following my investigation, it seems reasonable to accept the location data for these records and correct the north latitude point to be -31.92 instead.

Flipping coordinate:
```{r}
decipiens_clean <- decipiens_clean %>% 
  mutate(decimalLatitude = if_else(decimalLatitude > 0, -decimalLatitude, decimalLatitude))
```

Note in this case that was the only north latitude point so I don't need to worry about the code incorrectly flipping other points' coordinates.

Remake the map to check:

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F")
```
Success!

Next up I want to investigate the observations on Australia's East coast which seem very out of place.

Looking at observations with high longitude, they are all super old (1960-1971) records from AD AVH Data and Canberra. They are also preserved specimens. 

### Removing Preserved Specimens

After chatting with Rob, I've confirmed that preserved specimens should be removed since they're not a true indication of where E. decipiens naturally occur.

This makes it easier, as many of the problematic-looking points seem to be preserved specimens.
```{r}
nrow(decipiens_clean)
```

```{r}
decipiens_clean <- decipiens_clean %>% 
  filter(!basisOfRecord == 'PRESERVED_SPECIMEN')
```

```{r}
nrow(decipiens_clean)
```
Wow! That removed a whopping
```{r}
551-185
```
observations! Now viewing the map again:

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.4)
```

Now the IUCN expert range is starting to look a lot more apt than it did before!
Let's zoom in to the relevant area a bit more.
```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.4) +
  coord_sf(xlim = c(114, 120),
           ylim = c(-36, -29),
           expand = FALSE)
# Zoom code made with help of ChatGPT:
# https://chatgpt.com/share/6a066eec-e304-83ec-b523-cf49140ca891
```

Note that in the top tip of the IUCN expert range, there was a larger cluster of points around the 2 observations that remain there in my cleaned dataset. Was there some reason for the IUCN to include more of them? Well, looking at the full dataset prior to cleaning, almost all points in that area (around 30S, 116E) were preserved specimens, except the exact 2 that I have left. For this reason I'll stand by my removal decision.

I'm now going to remake the ggpointdensity plot to see how well overlapping observations have been dealt with after some cleaning:

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens_clean,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) + 
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F", alpha = 0.2) +
  coord_sf(xlim = c(114, 120),
           ylim = c(-36, -29),
           expand = FALSE) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```

Looking on Google Maps, the large cluster is indeed around Perth, especially north of Perth where there is more forest and less urban sprawl. For example the Gangara-Moore River State forest has coords around -31.5, 116 which lines up with high clustering of E. decipiens occurrence records. 

https://www.google.com/maps/@-31.5593691,115.8296231,25841m/data=!3m1!1e3?entry=ttu&g_ep=EgoyMDI2MDUxMi4wIKXMDSoASAFQAw%3D%3D


I also noticed the other, less prominent cluster has many points in and around Stirling National Park.

https://www.google.com/maps/@-34.3899856,117.3228533,196844m/data=!3m1!1e3?entry=ttu&g_ep=EgoyMDI2MDUxMi4wIKXMDSoASAFQAw%3D%3D

It is my suspicion that there are far more E. decipiens out there in the other forest areas that are less visited by humans, and these clusters are primarily the result of bias towards more populated/visited areas, parks and roads.

Once I've made my decisions on how to clean the data and remade this plot, 
I can:

# Assess whether expert range map should be changed

# Build SDM for E. decipiens suitable climate space

* look at soil and landscape grid of Australia, in addition to CHELSA and WorldClim etc.
* Add soil atlas into the model
