---
title: "Data Science for Biologists Final Project: EDA"
author: "Jack Davies u7921974"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

## Location on Github

https://github.com/anu-jackdavies/DS4B-final-project


#### My Eucalypt Species - Eucalyptus decipiens

# Download ALA data and IUCN map

```{r}
library(galah)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(readxl)
library(janitor)
library(purrr)
library(sf)  
library(ozmaps)
library(here) 
```

Using galah_call to get ALA occurrence data for **Eucalyptus decipens**:
```{r}
galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_counts()
```

```{r}
galah_config(email = "iamjackdavies@gmail.com")
decipiens <- galah_call() %>% 
  identify("Eucalyptus decipiens") %>% 
  atlas_occurrences()
```
Importing IUCN expert range map data for E. decipiens:
```{r}
IUCN_map <- st_read("raw_data/Eucalyptus_decipiens.gpkg")
```
Now 'IUCN_map' can be used to add the spatial polygon 

# Clean occurrence data

This will be a key part of the project which needs much consideration.

Firstly, I'll plot all of the occurrence data onto a map, then overlay the expert range.
This will be a good way to visualise what the spatial distribution of E. decipiens looks like.

## Plot ALA occurrence data

```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3')
```
Could potentially flip the point with latitude around +33 since if it were -33 it would align with many other points - look into this.


### Overlapping observations

Trying to visualise overlapping observations like in Week 9 workshop:
```{r}
library(ggpointdensity)
library(viridis)
```

```{r}
ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_pointdensity(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude)) +
  scale_color_viridis_c(direction = -1) +
  guides(colour = guide_colorbar(title = "Number of overlapping observations"))
```

## Overlay IUCN expert range on occurrence map
```{r}
aus <- ozmaps::ozmap_country

ggplot() + 
  geom_sf(data = aus,
          fill = "white",
          colour = "grey60") +
  geom_point(data = decipiens,
             aes(x = decimalLongitude,
             y = decimalLatitude),
             colour = 'olivedrab3') +
  geom_sf(data = IUCN_map, 
          fill = "#009E73", color = "#007F5F")
```

Once I've made my decisions on how to clean the data and remade this plot, 
I can:

# Assess whether expert range map should be changed

# Build SDM for E. decipiens suitable climate space